Seeing Double: Transient Dips and Photometric Binarity in PTFO$\,$8-8695

### 2020-04-23 paper/ms.tex
astrometric residual: added the variability check

% \begin{figure*}[t]
% 	\begin{center}
% 		\leavevmode
% 		\includegraphics[width=0.9\textwidth]{f6.png}
% 	\end{center}
% 	\vspace{-0.7cm}
% 	\caption{ {\bf Positions, kinematics, and HR diagram of \ptfo.}
%   Members of the 25$\,$Ori-1 subgroup are shown with black circles,
%   and were identified by \citet{kounkel_apogee2_2018} through
%   clustering on Gaia DR2 and APOGEE data.  The ``neighborhood'' (gray
%   circles) is defined as the group of at most $10^4$ randomly selected
%   non-member stars within 5 standard deviations of the mean right
%   ascension, declination, and parallax.
% 	\label{fig:gaia}
% 	}
% \end{figure*}

@@

		\vspace{-0.7cm}
		\subfloat{
			\includegraphics[width=0.7\textwidth]{f5b.pdf}
		}
	\end{center}
	\vspace{-0.7cm}
	\caption{ {\bf Evidence for binarity in \ptfo}.
    {\it Top: Hertzsprung-Russell diagram of \ptfo\ and late-type
    members of 25$\,$Ori-1.} Members of the 25$\,$Ori-1 group (black
    circles) were identified by \citet{kounkel_apogee2_2018} through
    clustering on six-dimensional Gaia DR2 and APOGEE-2 data.  The
    ``neighborhood'' (gray circles) is non-member stars within 5
    standard deviations of the mean 25$\,$Ori-1 right ascension,
    declination, and parallax.  It contains members of the Orion
    complex with its full spread of ages, in addition to field
    interlopers.  $G$ denotes Gaia broadband magnitudes, $Bp$ Gaia
    blue, $Rp$ Gaia red, and $\omega_{\rm as}$ the parallax in
    arcseconds.  The $x$-axis limits have been set to show only K and
    M dwarf members, to accentuate \ptfo's separation from the
    single-star sequence.  {\it Bottom: Astrometric goodness-of-fit
    versus $Rp$ magnitude for 25$\,$Ori-1 members}.  The single-source
    astrometric model provides a poor fit, which could be due to
    stellar variability or binarity.
		\label{fig:gaia}
	}
\end{figure*}

@@

A separate possible line of evidence for binarity is the Gaia DR2
astrometry.  As noted in Section~\ref{sec:observations}, the Gaia DR2
solution for \ptfo\ shows a 10.3$\sigma$ astrometric excess.  This
astrometric excess indicates the degree to which a single-source model
fails to explain the observed astrometric measurements.  Specifically,
the single-source astrometric model yielded $\chi^2=325.2$.  There are
121 astrometric measurements, and 5 free parameters, and therefore 116
degrees of freedom. The reduced $\chi^2$ is 2.80.  The majority of
stars with comparable brightness in Gaia do not show such poor
goodness-of-fit \citep[][Appendix A]{lindegren_gaiasoln_2018}.

@@

Using the same 149 members in the 25$\,$Ori-1 subgroup from
\citet{kounkel_apogee2_2018}, we calculated the reduced $\chi^2$ for
each member.  The lower panel of Figure~\ref{fig:gaia} shows the
resulting values, as a function of stellar brightness.  \ptfo\ is in
the upper 90$^{\rm th}$ percentile of stars showing astrometric
excesses within the 25$\,$Ori-1 group.  Relative to other M-dwarf
group members with comparable brightnesses and variability
characteristics, \ptfo\ still stands out as behaving astrometrically
poorly.  Ultimately, we will have to wait for the full release of the
nominal Gaia mission to definitively determine whether the astrometric
excess is caused by stellar binarity or photometric variability.
Nonetheless this result suggests that stellar binarity is indeed the
root cause, because other member stars have identical variability
characteristics but do not show such a large astrometric excess.

@@

\begin{figure}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.5\textwidth]{f6.pdf}
	\end{center}
	\vspace{-0.7cm}
	\caption{
		{\bf Timing residuals for \ptfob\ from a decade of monitoring.}
		Black points are times of dips, minus the indicated linear
		ephemeris.  The $y$-axis is given in units of phase for the
		short-period signal.  The star shows the binned TESS ephemeris.
		Dips have been observed by \citet{van_eyken_ptf_2012},
		\citet{ciardi_followup_2015}, \citet{yu_tests_2015},
		\citet{raetz_yeti_2016}, \citet{onitsuka_multicolor_2017}, and
		\citet{tanimoto_evidence_2020}.  Certain dips ({\it e.g.}, the one
		at phase 0 in mid-2019) are consistent with noise, and were likely
		reported because something was {\it expected}, rather than
		convincingly {\it observed}.  Horizontal dashed lines are drawn at
		$\pm (P_{\rm \ell} - P_{\rm s})/P_{\rm s}$, highlighting either a
		numerical coincidence or an observational bias.  The orbital phase
		observed by TESS is consistent with that of
		\citet{tanimoto_evidence_2020}, and quite different from the
		original phase.
		\label{fig:o_minus_c}
	}
\end{figure}

>>>>

@@

		\vspace{-0.7cm}
		\subfloat{
			\includegraphics[width=0.7\textwidth]{f5b.pdf}
		}
	\end{center}
	\vspace{-0.7cm}
	\caption{ {\bf Evidence for binarity in \ptfo}.
    {\it Top: Hertzsprung-Russell diagram of \ptfo\ and late-type
    members of 25$\,$Ori-1.} Members of the 25$\,$Ori-1 group (black
    circles) were identified by \citet{kounkel_apogee2_2018} through
    clustering on six-dimensional Gaia DR2 and APOGEE-2 data.  The
    ``neighborhood'' (gray circles) is non-member stars within 5
    standard deviations of the mean 25$\,$Ori-1 right ascension,
    declination, and parallax.  It contains members of the Orion
    complex with its full spread of ages, in addition to field
    interlopers.  $G$ denotes Gaia broadband magnitudes, $Bp$ Gaia
    blue, $Rp$ Gaia red, and $\omega_{\rm as}$ the parallax in
    arcseconds.  The $x$-axis limits have been set to show only K and
    M dwarf members, to accentuate \ptfo's separation from the
    single-star sequence.  {\it Bottom: Astrometric goodness-of-fit
    versus $Rp$ magnitude for 25$\,$Ori-1 members}.  The single-source
    astrometric model for \ptfo\ provides a poor fit, which could be due to
    stellar variability or binarity.  Cluster members that
    are at least as variable as \ptfo\ show lower astrometric excesses
    (black squares), suggesting binarity as the root cause.
		\label{fig:gaia}
	}
\end{figure*}

@@

A separate possible line of evidence for binarity is the Gaia DR2
astrometry.  As noted in Section~\ref{sec:observations}, the Gaia DR2
solution for \ptfo\ shows a 10.3$\sigma$ astrometric excess.  This
astrometric excess indicates the degree to which a single-source model
fails to explain the observed astrometric measurements.  Specifically,
the single-source astrometric model yielded $\chi^2=325.2$.  There are
121 astrometric measurements, and 5 free parameters, and therefore 116
degrees of freedom. The reduced $\chi^2$ is 2.80.  The majority of
stars with comparable brightness in Gaia do not show such poor
goodness-of-fit \citep[see][Appendix A]{lindegren_gaiasoln_2018}.

@@

% NB: CDIPS-I only made 131 light-curves for 25Ori-1, because I wasn't
% aware of the Kounkel+18 member list at the time.
Using the same 149 members in the 25$\,$Ori-1 subgroup, we calculated
the astrometric reduced $\chi^2$ for each member.  We then queried the
CDIPS lightcurve database at MAST \citep{bouma_cluster_2019} to find
the subset of stars that were at least as variable as \ptfo.
We measured the variability amplitude by taking the difference of the
$95^{\rm th}$ and $5^{\rm th}$ percentiles of the flux.
This yielded 30 stars of equal or greater variability.
The lower panel of Figure~\ref{fig:gaia} shows the
reduced $\chi^2$ as a function of stellar brightness.  \ptfo\ is in
the upper 90$^{\rm th}$ percentile of stars showing astrometric
excesses within the 25$\,$Ori-1 group.  Relative to other M-dwarf
group members with comparable brightnesses and variability
characteristics, \ptfo\ still stands out as behaving astrometrically
poorly.  Ultimately, we will have to wait for the full release of the
nominal Gaia mission to definitively determine whether the astrometric
excess is caused by stellar binarity or photometric variability.
Nonetheless the fact that comparably variable stars do not show
comparably large astrometric excesses suggests that stellar binarity
is indeed the root cause.

@@

\begin{figure}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.5\textwidth]{f6.pdf}
	\end{center}
	\vspace{-0.7cm}
	\caption{
		{\bf Timing residuals for \ptfob\ from a decade of monitoring.}
		Black points are times of dips, minus the indicated linear
		ephemeris.  The $y$-axis is given in units of phase for the
		short-period signal.  The star shows the binned TESS ephemeris.
		Dips have been observed by \citet{van_eyken_ptf_2012},
		\citet{ciardi_followup_2015}, \citet{yu_tests_2015},
		\citet{raetz_yeti_2016}, \citet{onitsuka_multicolor_2017}, and
		\citet{tanimoto_evidence_2020}.  Certain dips ({\it e.g.}, the one
		at phase 0 in mid-2019) are consistent with noise, and were likely
		reported because something was {\it expected}, rather than
		convincingly {\it observed}.  Horizontal dashed lines are drawn at
		$\pm (P_{\rm \ell} - P_{\rm s})/P_{\rm s}$, highlighting either a
		numerical coincidence or an observational bias.  The orbital phase
		observed by TESS (lower-right) is consistent with that of
		\citet{tanimoto_evidence_2020}, and quite different from the
		original phase.
		\label{fig:o_minus_c}
	}
\end{figure}
